Boosted Regression Trees for Data-Driven Auto-Tuning

### 2011-12-08 paper/draft1.tex
m

\documentclass{sig-alternate}
\usepackage{graphicx}
\usepackage{amsmath}
\usepackage{amssymb}

@@

\begin{abstract}
Auto-tuning is a widely used and effective technique for optimizing a
parametrized GPU code template for a particular computation on particular
hardware.  Its drawback is that thorough or exhaustive auto-tuning requires
compiling many kernels and calling each one many times; this process is slow.
Furthermore, library abstraction boundaries provide operations such as image
filtering and matrix multiplication, which actually correspond to a large set
of potential problem configurations with a wide variety of memory access
patterns and computational bottlenecks.  How can we draw on data from previous
empirical auto-tuning of related problems on related hardware to make a just-in-time
implementation decision for a novel problem?  This paper presents a machine
learning approach to auto-tuning, in which features of (a) the current
hardware platform, (b) the kernel configuration and (c) the problem instance
are passed to a regression model (boosted regression trees) which predicts how
much faster this kernel will be than a reference baseline.  Combinatorial
optimization strategies that would normally implement auto-tuning by
evaluating kernel configurations on the real hardware are orders of magnitude
faster when evaluating the regression model instead.  We validate our approach
using the filterbank correlation kernel described in [XXX], where we find
that 0.1 seconds (XXX) of hill climbing on the regression model can achieve an
average of 95\% of the speed brought by 120 seconds of true auto-tuning XXX.  Our
approach is not specific to filterbank correlation, or even to GPU
auto-tuning: the approach of using a non-linear regression model on top of
simple features applies to a variety of problem types, kernel types, and
platforms.
\end{abstract}

@@

Autotuning is a family of empirical techniques for finding the implementation
that minimizes that runtime function, when the platform and the problem
configuration are given.

@@

\section{Filter-bank Correlation}

@@

Filterbank correlation is used extensively in image and video processing,
where it is often the computational bottleneck.
GPUs are well-suited to computing filterbank correlations, although the
optimal strategy for doing so often depends on the shape of the problem, i.e.
the choices of $R$, $C$, $D$, $F$, and $W$.
The computation of filterbank correlation on GPUs was recently treated by
XXX, which serves as the basis for the present work.

\section{Problem Configuration}

We randomly sampled problem specifications from the following space:
\begin{align*}
R = C & \in \{ 256, 512, 1024, 2048, 4096 \} \\
H = W & \in \{ 3, 5, 7, 9, 11 \} \\
D &  \in \{1, 4, 8, 16, 32, 64, 128, 256 \} \\
F &  \in \{1, 4, 8, 16, 32, 64, 128, 256 \}
\end{align*}
The problem configuration space used in our experiments. The product space
includes 1600 elements, but we restricted our experiments to problems that
represented between 1 and 50 gigaflops. Our problem space included 602
configurations. In a full application in which all image and filter sizes are
valid, the number of configurations would be effectively infinite.

% Problems were considered invalid if the filters were larger than the images.
% but in the new setting with iheight >= 256 this never happened.
Problems were tested if they represented between 1 and 50 gigaflops.  Problems
that were smaller do not fully utilize the hardware, and are handled equally
well by many kernel settings, including the reference (XXX cite GCG).
Problems that were larger take so long to evaluate that there is negligible
inefficiency in implementing them via multiple calls with smaller images and fewer filters.
The configuration space described in Figure~\ref{fig:probspace} included 602
configurations with between 1 and 50 gigaflops.

In a full, reusable, library implementation of the filter correlation
operation, we would also provide options for arbitrary image shapes,
arbitrary integer depth and filter shapes, and also support a variety of
physical input layouts via strides.  These additional options would make our
approach of automatic auto-tuning even more important, because there would be
a greater variety in the kinds of computations and memory transfers to
perform.  In our experiments, we only consider the problem variations
described in Figure ~\ref{tbl:}, which already suffice to show that automatic
auto-tuning is a promising approach.

\subsection{Kernel Space: Metaprogramming}

\begin{align*}
\mathrm{block\_w}    & \in (4, 8, 16, 32, 64, 128) \\
\mathrm{block\_h}    & \in (4, 8, 16, 32, 64, 128) \\
\mathrm{n\_filter\_r} & \in (1, 2) \\
\mathrm{n\_output4s} & \in (\mathrm{all}, 1, 2) \\
\mathrm{spill}      & \in (False, True) \\
\mathrm{imul\_fast}  & \in (False, True) \\
\mathrm{pad\_shared} & \in (False, True) \\
\mathrm{use\_tex1d}  & \in (False, True) \\
\mathrm{maxrreg}    & \in (8, 16, 20, 24, 28, 32, \infty) \\
\mathrm{fast\_math}  & \in (False, True)
\end{align*}

@@

\iffalse
\subsection{Platform Space: XXX}

XXX: so far we only tried 3 cards... so we're using a 1-of-N feature.
\fi

\section{Results}

@@

\end{document}

>>>>

\documentclass{sig-alternate}
\usepackage{graphicx}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{natbib}
\usepackage{myAlgorithm}

\bibliographystyle{abbrvnat}

@@

\begin{abstract}
Auto-tuning is a widely used and effective technique for optimizing a
parametrized GPU code template for a particular computation on particular
hardware.  Its drawback is that thorough or exhaustive auto-tuning requires
compiling many kernels and calling each one many times; this process is slow.
Furthermore, library abstraction boundaries provide operations such as image
filtering and matrix multiplication, which actually correspond to a large set
of potential problem configurations with a wide variety of memory access
patterns and computational bottlenecks.  How can we draw on data from previous
empirical auto-tuning of related problems on related hardware to make a just-in-time
implementation decision for a novel problem?  This paper presents a machine
learning approach to auto-tuning, in which features of (a) the current
hardware platform, (b) the kernel configuration and (c) the problem instance
are passed to a regression model (boosted regression trees) which predicts how
much faster this kernel will be than a reference baseline.  Combinatorial
optimization strategies that would normally implement auto-tuning by
evaluating kernel configurations on the real hardware are orders of magnitude
faster when evaluating the regression model instead.  We validate our approach
using the filterbank correlation kernel described in \citet{pinto+cox:2011gcg}, where we find
that 0.1 seconds (XXX) of hill climbing on the regression model can achieve an
average of 95\% of the speed brought by 120 seconds of true auto-tuning XXX.  Our
approach is not specific to filterbank correlation, or even to GPU
auto-tuning: the approach of using a non-linear regression model on top of
simple features applies to a variety of problem types, kernel types, and
platforms.
\end{abstract}

@@

Auto-tuning is a family of empirical techniques for finding the implementation
that minimizes that runtime function, when the platform and the problem
configuration are given.

@@

\cite{volkov+demmel:2008}
\cite{vuduc:2003} % has some stuff about ML for autotuning in ch. 9

\subsection{Filter-bank Correlation}

@@

Filterbank correlation is used extensively in image and video processing,
where it is often the computational bottleneck.
GPUs are well-suited to computing filterbank correlations, although the
optimal strategy for doing so often depends on the shape of the problem, i.e.
the choices of $R$, $C$, $D$, $F$, and $W$.
The computation of filterbank correlation on GPUs was recently treated by
\citet{pinto+cox:2011gcg}, which serves as the basis for the present work.

XXX: TYPICAL OF STENCIL COMPUTATIONS.

``Although the auto-tuning strategy has been successfully applied to libraries,
generalized stencil kernels are not amenable to packaging as libraries.''
\cite{kamil+etal:2009}

\cite{datta:2009}

\section{Configuration Spaces}

Recall from the introduction (Eq.~\ref{eq:condopt}) that we are addressing a
conditional optimization problem, in which we condition on a problem
configuration ($P$) and a system specification ($S$), and search for the best
implementation $I$ for addressing problem $P$ on system $S$.
This section describes the configuration spaces we used in our tests,
which are described later in Section XXX.

\subsection{Problem Description Space}

Mathematically, filterbank correlation is an operation on an $R \times C \times
D$ image and a $F \times H \times W \times D$ filterbank.
We randomly sampled problem specifications from the following space:
\begin{align*}
R = C & \in \{ 256, 512, 1024, 2048, 4096 \} \\
H = W & \in \{ 3, 5, 7, 9, 11 \} \\
D &  \in \{1, 4, 8, 16, 32, 64, 128, 256 \} \\
F &  \in \{1, 4, 8, 16, 32, 64, 128, 256 \}
\end{align*}
The product space
includes 1600 elements, but we restricted our experiments to problems that
represented between 1 and 50 gigaflops. Our problem space included 602
configurations. A library implementation that must support all image and
filter sizes must handle on the order of $2^{50}$ forseable
problem configurations (not even including variations due to strided memory
layouts).

% Problems were considered invalid if the filters were larger than the images.
% but in the new setting with iheight >= 256 this never happened.
Problems were tested if they represented between 1 and 50 gigaflops.  Problems
that were smaller do not fully utilize the hardware, and are handled equally
well by many kernel settings.
Problems that were larger take so long to evaluate that there is negligible
inefficiency in implementing them via multiple calls with smaller images and fewer filters.
The configuration space described in Figure~\ref{fig:probspace} included 602
configurations with between 1 and 50 gigaflops.

In a full, reusable, library implementation of the filter correlation
operation, we would also provide options for arbitrary image shapes,
arbitrary integer depth and filter shapes, and also support a variety of
physical input layouts via strides.  These additional options would make our
approach of automatic auto-tuning even more important, because there would be
a greater variety in the kinds of computations and memory transfers to
perform.

\subsection{Implementation Description Space}

The strategy we use for computing filterbank correlation on the GPU
using CUDA follows \citet{pinto+cox:2011gcg}.
The overall strategy is to load the filterbank into some relatively fast
memory (constant or texture), and then launch a grid of blocks that each
computes some local piece of the output image $z$.
Each thread computes $4 \times n\_output\_4s$ channels for some column and row of $z$.
Each block computes a subrectangle of the output $z$,
and the grid is used to tile the whole output image with blocks.
Some filterbanks are too large to fit into the fast memory.
For large filterbanks our approach permits splitting the filterbanks along
the number-of-filters dimension ($F$) and the number-of-rows dimension ($H$).
All the filterbanks in our study are small enough that at least half of each
filter can always fit into the constant or texture memory.
When splitting the filterbank in this way we must use several kernel
executions, each of which increments the output buffer by a partial result.
Pseudo-code for the kernel is given in Figure~\ref{fig:kernel}.

\begin{figure}[!ht]
\centerline{
\begin{algorithm}{$\Algo{fbcorr}\big(gX, cF, gZ \big)$}
\Aitem shared $sX \setto$ all channels of region ($\beta$) of $gX$
\Aitem $x, y \setto$ position of this thread in output image
\Aitem \_\_syncthreads()
\Aitem $v[0:N] \setto 0$, for $N=4\times n\_output\_4s$
\Aitem for $d \setto 0$~\To~$D$,
\Aitem ~ for $h \setto 0$~\To~$H / n\_filter\_r$,
\Aitem ~ ~ for $w \setto 0$~\To~$W$,
\Aitem ~ ~ ~ $u \setto sX[x+h, y+w, d]$
\Aitem ~ ~ ~ for $n \setto 0$~\To~$n\_output\_4s - 1$,
\Aitem ~ ~ ~ ~  $v[n] \setto v[n] + cF[n, h, w, d]$
\Aitem for $n \setto 0$~\To~$n\_output\_4s - 1$,
\Aitem ~  gZ[x][y][4n:4n+n] = v[4n:4n+n], (float4)
\end{algorithm}
}
\caption{Kernel pseudo-code for filterbank correlation.
Input $gX$ is a pointer to $x$ in global memory, input $cF$ is a pointer to
$f$ in either constant or texture memory, and output $gZ$ is a pointer to $z$
in global memory.
Each block of threads modifies $4 \times n\_output\_4s$ channels of a
rectangle (called $\beta$ in code listing) within $z$. A grid of blocks covers all rows and columns of $z$.
Multiple calls can be used to apply all filters of a large filterbank $f$ to
$x$.
}
\label{fig:kernel}
\end{figure}

\begin{align*}
\mathrm{block\_w}    & \in (4, 8, 16, 32, 64, 128) \\
\mathrm{block\_h}    & \in (4, 8, 16, 32, 64, 128) \\
\mathrm{n\_filter\_r} & \in (1, 2) \\
\mathrm{n\_output\_4s} & \in (\mathrm{all}, 1, 2) \\
\mathrm{spill}      & \in (False, True) \\
\mathrm{imul\_fast}  & \in (False, True) \\
\mathrm{pad\_shared} & \in (False, True) \\
\mathrm{use\_tex1d}  & \in (False, True) \\
\mathrm{maxrreg}    & \in (8, 16, 20, 24, 28, 32, \infty) \\
\mathrm{fast\_math}  & \in (False, True)
\end{align*}

@@

Meta-programming was implemented in Python using
Cheetah for string processing~\citep{cheetah}
and PyCUDA~\citep{klochner+etal:2009} for dynamic kernel compilation.

XXX: analysis of READ traffic, WRITE traffic, DATA re-use and CACHEing
strategies. FLOPS per write etc.

XXX: Why not use FFT: convolution in spectral domain?

\subsection{System Description Space}

 XXXX Leave for future work?

\section{Auto-Tuning by Hill Climbing}

@@

Platform Space: XXX

XXX: so far we only tried 3 cards... so we're using a 1-of-N feature.

%\renewcommand{\bibsection}{\section*{Bibliography}}
%\setlength{\bibsep}{5pt}
\bibliography{local}
\end{document}
